### 2022-03-07 paper/localpsf.tex
paper edits

@@

Once the scattered impulse responses $\impulseresponse_{x_i}$ are computed, we approximate the integral kernel $\Phi(y,x)$ at arbitrary points $(y,x)$ by interpolation. The details of the interpolation will be described in Section SECTION, but the key principle behind the interpolation is an assumption that we call \emph{local mean-displacement invariance}. Specifically, we make the assumption that
\begin{equation}
\Phi(y,x) = \phi_x(y) \approx \phi_{x_i}(y - \mu(x) + \mu(x_i))
\end{equation}
if $x_i$ is close to $x$. This is shown in Figure FIG. We approximate $\phi_x(y)$ by a weighted linear combination of the values $\phi_{x_i}(y - \mu(x) + \mu(x_i))$ for the a small number of sample points $x_i$ near $x$. The weights in the linear combination are determined via radial basis function interpolation.

@@

Often we are interested in approximating operators $\Aop$ that are symmetric, or symmetric positive semi-definite. If $\Aop$ is symmetric, impulse responses of $\Aop$ are also impulse responses of $\Aop^T$, which may be interpolated in the variable $y$ instead of $x$ to yield a different approximation of $\Phi(y,x)$. In this case, we combine information from impulse responses of both $\Aop$ and $\Aop^T$ in the interpolation procedure for approximating $\Phi(y,x)$, to get a more accurate approximation at negligible additional cost. Intuitively, impulse responses of $\Aop$ can be thought of as ``columns'' of $\Phi$, impulse responses of $\Aop^T$ can be thought of as ``rows'' of $\Phi$, and $\Phi(y,x)$ can be thought of as an unknown entry of $\Phi$. When $\Aop$ is symmetric, computing a column of $\Aop$ also gives us a row of $\Aop$ for free, so we approximate $\Phi(y,x)$ using information from a combination of known columns and known rows, rather than just known columns.

When $\Aop$ is symmetric positive semi-definite, $\mathbf{A}_H$ may be indefinite due to errors in the approximation. In this case, we modify the H-matrix representation of $\mathbf{A}_H$ to make it positive semi-definite, by applying a specially chosen rational matrix function to $\mathbf{A}_H$.

@@

Let $x$ and $x'$ be points in $\Omega$ that are close to each other, and consider the following three approximations, ordered by increasing level of sophistication:
\begin{align}
	\phi_x(y) &\approx \phi_{x'}(y) \label{eq:translate1}\\
	\phi_x(y) &\approx \phi_{x'}(y-x+x') \label{eq:translate2}\\
	\phi_x(y) &\approx \phi_{x'}\left(y-\mu(x)+\mu(x')\right). \label{eq:translate3}
\end{align}
We use approximation \eqref{eq:translate3}, which we call ``locally mean-displacement invariance.'' This is illustrated in Figure \ref{fig:mean_displacement_invariance}.

Approximation \eqref{eq:translate1} follows from the assumption that $\Phi$ is smooth. Operators satisfying \eqref{eq:translate1} can be well-approximated via low-rank interpolative factorization of $\Phi$.
% of the form
%\begin{equation}
%	\label{eq:low_rank_interpolation}
%	\Phi(y,x) \approx \sum_{i=1}^r w_i(x) \phi_{x_i}(y)
%\end{equation}
%where $\phi_{x_i}$ are known impulse responses at a collection of points $x_i$ scattered throughout $\Omega$, and $w_i$ are smooth weighting functions that interpolate the $\phi_{x_i}$. 
%In problems of practical interest, \eqref{eq:translate1} typically holds \emph{asymptotically} in the limit as $x'$ approaches $x$. 
However, even if $\Phi$ is smooth (which is typically the case asymptotically) the relevant Lipschitz constant for continuity, and hence the required rank in the low rank approximation, can be large, which makes algorithms based on assumption \eqref{eq:translate1} computationally expensive. This is the case, for example, for Hessians in PDE constrained inverse problems with highly informative data. As the data become more informative, typically the ``mass'' of $\Phi(y,x)$ concentrates nearer and nearer to the ``diagonal'' $y=x$, the Lipschitz constant increases, and the required rank in the approximation increases.
%Moreover, if $\Aop$ is numerically low rank, then one would typically use more accurate and robust matrix-free methods for forming low-rank approximations, such as randomized SVD or Lanczos.

We use approximation \eqref{eq:translate3}, which is an improvement upon \eqref{eq:translate1} and \eqref{eq:translate2} that generalizes both. On one hand, if \eqref{eq:translate1} holds, then $\mu(x) \approx \mu(x_i)$, and so \eqref{eq:translate3} holds. On the other hand, translating a distribution translates its mean, so if \eqref{eq:translate2} holds, then $\mu(x_i)-\mu(x) \approx x_i - x$, so again \eqref{eq:translate3} holds. Furthermore, approximation \eqref{eq:translate3} can hold in situations where neither \eqref{eq:translate1} nor \eqref{eq:translate2} holds. For example, because the expected value is a linear operator, \eqref{eq:translate3} will hold when an $\Aop$ is locally translation-invariant with respect to a rotated frame of reference, while \eqref{eq:translate2} will not hold in this case. Additionally, \eqref{eq:translate3} generalizes to operators $\Aop:L^2(\Omega_1) \rightarrow L^2(\Omega_2)'$ that map between function spaces on different domains $\Omega_1$ and $\Omega_2$, and even operators that map between domains with different spatial dimensions. In contrast, \eqref{eq:translate2} does not naturally generalize to operators that map between function spaces on different domains, because the formula $y-x+x'$ requires $y \in \Omega_2$ and $-x+x' \in \Omega_1$ to be added together.

@@

\section{Method}
\label{sec:method}
In this section we describe how we compute the impulse response moments $V,\mu,\Sigma$ and use these moments to estimate the supports of the impulse responses $\phi_x$ (Section \ref{eq:mean_and_covariance_estimation}), the greedy ellipsoid packing procedure we use to choose batches of sample points $x_i$ (Section \ref{sec:sample_point_selection}), how we compute impulse responses, $\phi_{x_i}$, in batches (Section \ref{sec:get_impulse_response}), how we interpolate the $\phi_{x_i}$ to approximate integral kernel entries $\Phi(y,x)$ (Section \ref{sec:approximate_kernel_entries}), finite element discretization of $\Phi$ (Section \ref{sec:finite_element_kernel}), and how we construct the H-matrix approximation $\mathbf{A}_h$ of the operator $\mathcal{A}$. We also describe how to (optionally) modify $\mathbf{A}_h$ to force it to be symmetric positive semi-definite. The complete algorithm is shown in Algorithm \ref{alg:construct_Atilde}.

@@

\section{Impulse response moments}

@@

The idea here is that $\Phi(y,x) \approx f_i$ when $x_i$ is near $x$, so we approximate $\Phi(y,x)$ by interpolating the (point,value) pairs
\begin{equation}
\left(x_i, f_i\right), \qquad i=1,\dots,k,
\end{equation}
at the point $x$. To see that $\Phi(y,x) \approx f_i$, recall from Section \ref{eq:local_mean_displacement_invariance} that $\Phi(y,x) \approx \phi_{x_i}\left(y - \mu(x) + \mu(x_i)\right)$ if $x_i$ is near $x$. Inclusion of the extra factor $\frac{V(x)}{V(x_i)}$ in \eqref{eq:fxyxp} is a form of preconditioning that makes this approximation more robust in the case where $V(x)$ varies widely. 
%Note that we are not in danger of dividing by zero, because our ellipsoid packing procedure from Section \ref{sec:sample_point_selection} excludes $x_i$ from consideration as a sample point if $V(x_i)$ is smaller than a predetermined threshold.

>>>>

Let $x$ and $x'$ be points in $\Omega$ that are ``close'' to each other, and consider the following three approximations, ordered by increasing level of sophistication:
\begin{align}
\phi_x(y) &\approx \phi_{x'}(y) \label{eq:translate1}\\
\phi_x(y) &\approx \phi_{x'}(y-x+x') \label{eq:translate2}\\
\phi_x(y) &\approx \phi_{x'}\left(y-\mu(x)+\mu(x')\right). \label{eq:translate3}
\end{align}
Approximation \eqref{eq:translate1} says that $\phi_x$ can be approximated by $\phi_{x'}$ when $x$ and $x'$ are close. 
%This approximation follows from the assumption that $\Phi$ is smooth. 
Operators satisfying \eqref{eq:translate1} can be well-approximated via low-rank interpolative factorization of $\Phi$.
% of the form
%\begin{equation}
%	\label{eq:low_rank_interpolation}
%	\Phi(y,x) \approx \sum_{i=1}^r w_i(x) \phi_{x_i}(y)
%\end{equation}
%where $\phi_{x_i}$ are known impulse responses at a collection of points $x_i$ scattered throughout $\Omega$, and $w_i$ are smooth weighting functions that interpolate the $\phi_{x_i}$. 
%In problems of practical interest, \eqref{eq:translate1} typically holds \emph{asymptotically} in the limit as $x'$ approaches $x$. 
However, 
%even if $\Phi$ is smooth (which is typically the case asymptotically),
%the relevant Lipschitz constant for continuity, and hence 
the required rank in the low rank approximation can be large, which makes algorithms based on assumption \eqref{eq:translate1} computationally expensive. This is the case, for example, for Hessians in PDE constrained inverse problems with highly informative data, as discussed in Section \ref{sec:PDE_hessian_motivation}. 
%As the data become more informative, typically the ``mass'' of $\Phi(y,x)$ concentrates nearer and nearer to the ``diagonal'' $y=x$, the Lipschitz constant increases, and the required rank in the approximation increases.
%Moreover, if $\Aop$ is numerically low rank, then one would typically use more accurate and robust matrix-free methods for forming low-rank approximations, such as randomized SVD or Lanczos.

Our method is based on approximation \eqref{eq:translate3}, which is an improvement upon \eqref{eq:translate1} and \eqref{eq:translate2} that generalizes both. We call \eqref{eq:translate3} ``local mean-displacement invariance,'' and illustrate \eqref{eq:translate3} in Figure \ref{fig:mean_displacement_invariance}. On one hand, if \eqref{eq:translate1} holds, then $\mu(x) \approx \mu(x')$, and so \eqref{eq:translate3} holds. On the other hand, translating a distribution translates its mean, so if \eqref{eq:translate2} holds, then $\mu(x')-\mu(x) \approx x' - x$, so again \eqref{eq:translate3} holds. But approximation \eqref{eq:translate3} can hold in situations where neither \eqref{eq:translate1} nor \eqref{eq:translate2} holds. For example, because the expected value is a linear operator, \eqref{eq:translate3} will hold when an $\Aop$ is locally translation-invariant with respect to a rotated frame of reference, while \eqref{eq:translate2} will not hold in this case. Additionally, \eqref{eq:translate3} generalizes to operators $\Aop:L^2(\Omega_1) \rightarrow L^2(\Omega_2)'$ that map between function spaces on different domains $\Omega_1$ and $\Omega_2$, and even operators that map between domains with different spatial dimensions. In contrast, \eqref{eq:translate2} does not naturally generalize to operators that map between function spaces on different domains, because the formula $y-x+x'$ requires vectors in $\Omega_2$ and $\Omega_1$ to be added together.

@@

Once the scattered impulse responses $\impulseresponse_{x_i}$ are computed, we approximate the integral kernel $\Phi(y,x)$ at arbitrary points $(y,x)$ by interpolation. The details of the interpolation will be described in Section \ref{sec:approximate_kernel_entries}, but the key principle behind the interpolation is the local mean displacement invariance assumption discussed in Section \ref{eq:local_mean_displacement_invariance}.
%Specifically, we make the assumption that
%\begin{equation}
%\Phi(y,x) = \phi_x(y) \approx \phi_{x_i}(y - \mu(x) + \mu(x_i))
%\end{equation}
%if $x_i$ is close to $x$. This is shown in Figure \ref{fig:mean_displacement_invariance}. Motivation for this assumption is discussed in Section \ref{eq:local_mean_displacement_invariance}. 
We approximate $\Phi(y,x) = \phi_x(y)$ by a weighted linear combination of the values $\phi_{x_i}(y - \mu(x) + \mu(x_i))$ for a collection of sample points $x_i$ near $x$. The weights in the linear combination are determined via preconditioned radial basis function interpolation.

@@

%Often we are interested in approximating operators $\Aop$ that are symmetric, or symmetric positive semi-definite.

If $\Aop$ is symmetric, we take advantage of symmetry to construct a more accurate approximation at negligible additional cost. In this case, 
%impulse responses of $\Aop$ are also impulse responses of $\Aop^T$, which may be interpolated in the variable $y$ instead of $x$ to yield a different approximation of $\Phi(y,x)$. 
we combine information from impulse responses of both $\Aop$ and $\Aop^T$ in the interpolation procedure.
% for approximating $\Phi(y,x)$, to get a more accurate approximation at negligible additional cost.
Intuitively, impulse responses of $\Aop$ can be thought of as ``columns'' of $\Phi$, impulse responses of $\Aop^T$ can be thought of as ``rows'' of $\Phi$, and $\Phi(y,x)$ can be thought of as an unknown entry of $\Phi$. Therefore, when $\Aop$ is symmetric, computing a column of $\Aop$ also gives us a row of $\Aop$ for free, so we approximate $\Phi(y,x)$ using information from a combination of known columns and rows, rather than just known columns.

When $\Aop$ is positive semi-definite, $\mathbf{A}_H$ may be indefinite due to errors in the approximation. In this case, we modify the H-matrix representation of $\mathbf{A}_H$ to make it positive semi-definite, by applying a specially chosen rational matrix function to $\mathbf{A}_H$. This is described in Appendix \ref{sec:make_spd}.

@@

@@

\section{Method}
\label{sec:method}
In this section we describe how we compute the impulse response moments $V,\mu,\Sigma$ and use these moments to estimate the supports of the impulse responses $\phi_x$ (Section \ref{sec:moments}), the greedy ellipsoid packing procedure we use to choose batches of sample points $x_i$ (Section \ref{sec:sample_point_selection}), how we compute impulse responses, $\phi_{x_i}$, in batches (Section \ref{sec:get_impulse_response}), how we interpolate the $\phi_{x_i}$ to approximate integral kernel entries $\Phi(y,x)$ (Section \ref{sec:approximate_kernel_entries}), finite element discretization of $\Phi$ (Section \ref{sec:finite_element_kernel}), and how we construct the H-matrix approximation $\mathbf{A}_h$ of the operator $\mathcal{A}$. We also describe how to (optionally) modify $\mathbf{A}_h$ to force it to be symmetric positive semi-definite. The complete algorithm is shown in Algorithm \ref{alg:construct_Atilde}.

@@

\section{Impulse response moments}
\label{sec:moments}

@@

The idea here is that $\Phi(y,x) \approx f_i$ when $x_i$ is near $x$, so we approximate $\Phi(y,x)$ by interpolating the (point,value) pairs
\begin{equation}
\left(x_i, f_i\right), \qquad i=1,\dots,k,
\end{equation}
at the point $x$. To see that $\Phi(y,x) \approx f_i$, recall from Section \ref{eq:local_mean_displacement_invariance} that $\Phi(y,x) = \phi_x(y) \approx \phi_{x_i}\left(y - \mu(x) + \mu(x_i)\right)$ if $x_i$ is near $x$. Inclusion of the extra factor $\frac{V(x)}{V(x_i)}$ in \eqref{eq:fxyxp} is a form of preconditioning that makes this approximation more robust in the case where $V(x)$ varies widely. 
%Note that we are not in danger of dividing by zero, because our ellipsoid packing procedure from Section \ref{sec:sample_point_selection} excludes $x_i$ from consideration as a sample point if $V(x_i)$ is smaller than a predetermined threshold.
